Debugging by \textit{lastChange}

### 2010-10-13 papers/pldi2011/sigplanconf-template.tex

\begin{abstract}
Backward search from bug symptoms to defects-which cause the bug-is a natural way to debugging. However, due to the lack of support for easily bakward movement by  traditional debugging approaches (e.g., breakpoint-based or log-based debugging), applying this strategy-if it is not pracatically impossible-demands a lot of effort from developers. A fundumental step in this process is tracking origins to wrong values and any aid in this regard can greatly reduce the complexity of debugging for developers.

In this paper we present a new functionality in debuggers, \textit{lastChange}, which locates the last place a value is changed and lets developer to collect data and ask for more \textit{lastChange} queries at the point. The only prerequisite for \textit{lastChange} functionality is bug reproduciblity (i.e., there is a test case which reproduces the bug). We explain the algorithms behind this functionality.

As a proof of concept we developed \textit{RevNav}, a JavaScript prototype, which adds the \textit{lastChange} feature to Firebug JavaScript debugger. Moreover, \textit{RevNav} provides mechanism for automated execution reproduction, and a centralized user interface for all collected results.

@@

%---------------------------------------------------------------------------------------------------
\section{Introduction}
Debugging is an inevitable part of programming, still hard and time-consuming. To fix a bug, developers have to reproduce and monitor the buggy execution several times to understand the program's unexpected behavior. Trial and error, guess-work and analyzing huge collected data are the inseparable parts of this process. According to \cite{LaToza}, developers spend about fifty percent of their time debugging. Therefore, every improvement to debugging techniques and tools can considerably save developers' time and improve programs' quiality.

@@

Our proposal is a new functionality in debuggers which locates the origin of a wrong value, call it \textit{lastChange}. The only prerequisite for \textit{lastChange} functionality is bug reproduciblity (i.e., there is a test case which reproduces the bug). Assume that the program execution is paused on a breakpoint and the developer is suspicious about the value of an object property or variable. The developer selects \textit{lastChange} on the value and it is due to debugger to locate and shows the last change of object property or variable. Debugger reproduces the buggy execution and collects limited data and once the execution reaches the same place (i.e., the same breakpoint hit), call it \textit{reproduction point}, it pauses the execution, analyzes the collected data and shows the location of the last change to the developer. Developer can also examine the program state at the located point of execution, and asks for a new \textit{lastChange} at this point.

@@

In contrast, we propose a high-level function in debugger, \textit{lastChange}, which provides the answer without demanding anymore effort from the developer. By right clicking on \texttt{myObject.myProperty} in the right panel, the developer can run \textit{lastChange} command (Figure ~\ref{fig:example2}). Debugger considers the paused point at the breakpoint as the \textit{reproduction point}, and re-executes the buggy execution, pauses at the reproduction point and shows line 29 in a new panel (Figure ~\ref{fig:example3}). The program state debugger shows on the right panel belongs to the past. Therefore the right panel just shows the partially collected data. If the developer needs some values which are not available in the right panel, debugger re-executes another time and collects and shows the requested data.

@@

%---------------------------------------------------------------------------------------------------
\section{\textit{lastChange} Algorithm}
\textit{lastChange} can be called on two different types of values: An object property or a variable value. A variable can be a global, local or closure variable. We explain each case separately.

\subsection{\textit{lastChange} on object propery}

At the beginning, we assume that every object has a unique id. Moreover, we assume that there is a special watchpoint, \textit{property watchpoint}, that can be set on a property name \texttt{foo} and whenever property \texttt{foo} in any object changes this watchpoint hits. We know that none of these assumptions is true in JavaScript. We discuss this issue later in tracking object creation.

Once developer asks for the last change of \texttt{bar.foo} at a reproduction point, debugger sets a property watchpoint on \texttt{foo} and  re-executes the program. Whenever the property watchpoint hits, debugger keeps the object id and stack frame locations. Whenever the execution reaches the reproduction point it looks at the list of all changes and finds the last \texttt{foo} change with the same object id as \texttt{bar} id at the reproduction point(Figure ~\ref{fig:foo-changes1}).

>>>>

\begin{abstract}
Backward search from bug symptoms to defects-which cause the bug-is a natural way to debugging. However, due to the lack of support for easily bakward movement by  traditional debugging approaches (e.g., breakpoint-based or log-based debugging), applying this strategy-if it is not pracatically impossible-demands a lot of effort from developers. Tracking origins to wrong values is a fundumental step in this process and any aid in this regard can greatly reduce the complexity of debugging for developers.

To attack the mentioned problem, we propose a new functionality in debuggers, \textit{lastChange}, which locates the last place a value is changed and lets developer to collect data and ask for more \textit{lastChange} queries at the located point. While \textit{lastChange} algorithm is based on program re-execution, contrary to other replay-based algorithms which require exactly the same re-executions, it only requires \textit{bug reproduciblity} (i.e., a test case is available which reproduces the bug).

As a proof of concept we developed \textit{RevNav}, a JavaScript prototype, which adds the \textit{lastChange} feature to Firebug JavaScript debugger. Moreover, \textit{RevNav} provides mechanism for automated bug reproduction, and a centralized user interface which summarizes investigated execution points and collected results.

@@

%---------------------------------------------------------------------------------------------------
\section{Introduction}
Debugging is an inevitable part of programming, still hard and time-consuming. To fix a bug, developers have to reproduce and monitor the buggy execution several times to understand the program's unexpected behavior. Trial and error, guess-work and analyzing huge collected data are the inseparable parts of this process. According to \cite{LaToza}, developers spend about fifty percent of their time debugging. Therefore, every improvement to debugging techniques and tools can considerably save developers' time and improve programs' quality.

@@

Our proposal is a new functionality in debuggers which locates the origin of a wrong value, call it \textit{lastChange}. The only prerequisite for \textit{lastChange} functionality is bug reproduciblity (i.e., there is a test case which reproduces the bug). Assume that the program execution is paused on a breakpoint and the developer is suspicious about the value of an object property or variable. The developer selects \textit{lastChange} on the value and it is due to debugger to locate and shows the last change of object property or variable. Debugger reproduces the buggy execution and collects limited data and once the execution reaches the same place (i.e., the same breakpoint hit), call it \textit{reproduction point}, it pauses the execution, analyzes the collected data and shows the location of the last change to the developer. Developer can also examine the program state at the located point of execution, and ask for a new \textit{lastChange} at this point.

@@

In contrast, we propose a high-level function in debugger, \textit{lastChange}, which provides the answer without demanding anymore effort from the developer. By right clicking on \texttt{myObject.myProperty} in the right panel, the developer can run \textit{lastChange} command (Figure ~\ref{fig:example2}). Debugger considers the paused point at the breakpoint as the reproduction point, and re-executes the buggy execution, pauses at the reproduction point and shows line 29 in a new panel (Figure ~\ref{fig:example3}). The program state debugger shows on the right panel belongs to the past. Therefore the right panel just shows the partially collected data. If the developer needs some values which are not available in the right panel, debugger re-executes another time and collects and shows the requested data.

@@

%---------------------------------------------------------------------------------------------------
\section{\textit{lastChange} Algorithm}

The \textit{lastChange} algorithm is based on program re-execution. \textit{Reproduction point} is the execution point in which first \textit{lastChange} is calld. Like the introductory example, it can be specified by a [conditional] breakpoint. \textit{lastChange} can be called on two different types of values: An object property or a variable value. In JavaScript, every available variable in a frame comes from one of the scopes in the current scope chain. There are five different scopes: global, local, with, catch, closure. We explain each case separately.

\subsection{\textit{lastChange} on object propery}
To simplify the algorithm explanation and avoid technical details, we assume two basic operations and postpone implementation details to the next section. The first opertion is \texttt{objectId()} which gets an object and returns an integer as its identifier. This identifier is unique for every object during one execution. The second operation is \texttt{setPropertyChangeHook()} which gets function as the hook and a string as the property name like \texttt{foo}. Whenever property \texttt{foo} in any object changes the hook will be called. The hook function receives a reference to the owner of \texttt{foo}.

Once developer asks for the last change of \texttt{bar.foo} at the reproduction point, debugger calls \texttt{setPropertyChangeHook()} with \texttt{foo} as the property name and re-executes the program. Whenever \texttt{foo} changes and the hook function is called, debugger first calls \texttt{objectID()} on the \texttt{foo} owner object. Then it stores owner id and stack frame locations. Whenever the execution reaches the reproduction point it looks at the list of all changes and finds the last \texttt{foo} change with the same object id as \texttt{bar} id at the reproduction point(Figure ~\ref{fig:foo-changes1}).
